---
title: "Lab 1: exponential growth"
author: "NRES 470/670"
date: "Jan 24, 2017"
output: 
  html_document: 
    theme: spacelab
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```

In this lab we will have an opportunity to play around with the three main software packages we will use in this course: MS Excel, R, and InsightMaker. 

## Nomenclature for Population Ecology

First of all, we need a symbol to represent the population size. This is $N$!

$\Delta N$ represents the change in population size, ${N_{t+1}}-{N_t}$

The famous "BIDE" equation is a way to break down $\Delta N$ into components. 

$\Delta N = B + I - D - E \qquad \text{(Eq. 1)}$

where $B$ represents the number of births, $I$ reprents the number of immigrants, $D$ represents the number of deaths, and $E$ represents the number of emigrants. 

If we ignore immigration and emigratrion, then the BIDE equation simplifies to:

$\Delta N = B - D \qquad \text{(Eq. 2)}$

Now let's focus on $B$ and $D$. The important thing to recognize is that the number of births and deaths in a population is not constant. 

What does the number of births depend on? 

What **is** more likely to be constant is the per-capita rate of producing offspring, or dying. Does this make sense? These per-capita rates are often expressed as lower case letters. So $b$  represents per-capita births, and $d$ represents per-capita deaths. 

$b = \frac {B_t}{N_t} \qquad \text{(Eq. 3)}$

--or--

$B_t = b \cdot N_t$

The letter $t$ of course represents time, which could be years, or minutes, or decades, or whatever! So the above equation could be described as follows: "the number of births at a given time is equal to the per-capita birth rate times the total population size at that time"

Similarly, 

$D_t = d \cdot N_t \qquad \text{(Eq. 4)}$

Okay, we're almost there.

If $\Delta N = B - D \qquad \text{(Eq. 5)}$ 

then

$\Delta N = b \cdot N_t - d \cdot N_t\qquad \text{(Eq. 6)}$

which is equal to 

$\Delta N = (b - d) \cdot N_t \qquad \text{(Eq. 7)}$

which could also be written:

$\Delta N = r \cdot N_t\qquad \text{(Eq. 8)}$  

Where $r$ represents the difference between births and deaths. If $r$ is positive, then births are greater than deaths and the population grows. If $r$ is negative then deaths exceed births and the population declines. 

We can use *calculus notation* to consider the instantaneous change in population size:

$\frac{\partial N}{\partial t} = r \cdot N  \qquad \text{(Eq. 9)}$ 

This is probably the most fundamental equation of population ecology. 

If you integrate this equation, you get an equation that describes the population size at any time $t$:

$N_t = N_0 e^{rt}  \qquad \text{(Eq. 10)}$      

That is, population size at time $t$ is equal to the population size at time zero (initial abundance) multiplied by the root of the natural logarithm, $e$ to the $rt$ power. 

There you have it! Now you can compute population growth!

A couple more quick notes:

The greek symbol lambda ($\lambda$), is used to represent the finite rate of growth, or $\frac {N_{t+1}}{N_t}$. Lambda is what you multiply the current population size by to compute the population size in the next time step. Sometimes this is also called "big-R", or $R$. 

$N_{t+1}=N_t + B - D  \qquad \text{(Eq. 11)}$         

$N_{t+1}=N_t + b \cdot N_t - d \cdot N_t  \qquad \text{(Eq. 12)}$ 

$N_{t+1}=N_t + (b - d) \cdot N_t \qquad \text{(Eq. 13)}$       

$N_{t+1}=N_t + r \cdot N_t  \qquad \text{(Eq. 14)}$   

$N_{t+1}=N_t \cdot (1 + r)  \qquad \text{(Eq. 15)} $         

$N_{t+1}=\lambda \cdot N_t  \qquad \text{(Eq. 16)}$           


Okay now that we got that over with let's start the lab. The first software we will use is our old friend, Excel!

## Exponential growth in Excel

1.	Open the Excel spreadsheet [ExpGrowthExcel.xlsx](ExpGrowthExcel.xlsx). In the first column, we have a timestep of 1 year for 30 years. In the second column, we have an initial population size ($N_0$) of 100 individuals.  We also have a per-capita rate of increase ($r$) that is currently set at 0.1.


2.	To generate $N_t$ for the remaining time steps, we need to apply our knowledge of population ecology. Specifically we need to apply equation 15, above. Do this by clicking in the empty N(2) cell (position B3), typing =, clicking on N(1), typing the equation and hitting enter. As you do this, you should see the equation you are creating appear in the “functions bar”
 

3.	You can fill the remainder of the cells using the same equation for the other time steps by clicking and dragging the small square at the bottom of the N(2) cell that appears when the cell is selected.
 

4.	What happened?  We are not seeing a growing population here- actually it seems quite flat… this is surely not what we want!  Click on the N(3) cell to see what equation is being used to calculate the cell value.  The equation is B3*(1+D4).  The B3 part is correct – we want to calculate the N(3) population size using the $N$ from the previous timestep – but the D4 part is incorrect.  We always want to use the same $r$ which is always in the same cell.  You can see that when you drag down an equation as we have done, Excel adds 1 for each row so that the equation references the same relative positions in the spreadsheet for each new cell you want to calculate.  We like that Excel did that for $N$, but not for $r$, so we can tell Excel to stay in the same row for $r$ by inserting a dollar sign in our equation.

5.	In the N(2) cell, edit the equation in the “functions bar” so that there is a dollar sign in D3 (or just use the F4 shortcut). 
 
6.	Now drag the equation down again, and you should have a population size in row 31 of 1586.31

7.	Now we will plot our population against time.  Select both columns of data, and select the scatter plot under the Insert banner. A plot of N by Time will automatically appear.  You can change the $r$ value, the data and chart will automatically adjust.

### Exercise 1 (Excel-related problems)
Please provide short answers to the following questions, and provide your Excel spreadsheet to back up your answers.  

a. Apply equation 10 (above) to compute expected population size in year 30. Do you get the same answer? Why or why not?  What about year 100?  

b. What are the *units* of the per-capita rate of population growth, $r$? What if the time step were months instead of years? Would $r$ change? Try it! Use both methods (eq 15 vs equation 10) to compute population size in year 30. Do you see any difference? Is the difference than it was with a 1-year time step. Why or why not? 

c. What is the difference between continuous population growth and discrete population growth? Can you think of at least one case where continuous and discrete growth (respectively) would be a more biologically realistic model than the other? Explain.


## Exponential growth in R

R is probably the most common software used by modern ecologists for data analysis and simulation. There is a little bit of a learning curve with R, and I think InsightMaker is better in some ways for teaching this class, but I would be remiss if I didn't introduce you to R. We will do more with R when we get into data analysis!


### SET UP
Log onto computers using your assigned login and initial password (instructor will provide). Open the R software from the program menu or desktop.     

## PROCEDURE  
### STEP I: Set up R!
Go to website [http://cran.r-project.org/](http://cran.r-project.org/). This is the source for the free, public-domain R software and where you can access R packages, find help, access the user community, etc. The instructor will walk you through this website and provide some discussion on the R software program.

### STEP II. Take some time to get familiar with R
Take a quick look at the R manual, [Introduction to R](http://cran.r-project.org/doc/manuals/R-intro.html). To jump into the deep end of the pool, try to implement the steps in Appendix A, located [here](http://cran.r-project.org/doc/manuals/R-intro.html#A-sample-session). You might not understand everything right away, but you have the link, so you can return to this!  

If you already have basic R expertise, this is your opportunity to help your peers to develop the level of comfort and familiarity with R that they will need to perform data analysis and programming tasks in this course.     

Depending on whether you are already familiar with R, you may also find the remainder of this document useful as you work your way through the course (and there are many other good introductory R resources available online... let me know if there is one you particularly like and I will add it to the course website (Links page). As you work your way through this tutorial (on your own pace), please ask the instructor or your peers if you are uncertain about anything. 

### Set up the workspace

The first thing we usually do when we start an R session is we set up the workspace. That is, we load **packages**, assign key parameters and initialize variables. 

In this case, setting up the workspace is easy. We just need to define our parameter of interest - $r$ -, and set up a **vector** to represent the years of interest. 

We can store data in memory by assigning it to an "object" using the assignment operator **<-**. For example, this would assign the object "x" the value of 5. 

```{r eval=FALSE}
x <- 5
x
```

Let's assign our per-capita population growth rate, $r$ (but this could be called anything), and our initial population size to an object called **N0**, and the number of years to simulate.

```{r}
r <- 0.1
N0 <- 100
nyears <- 50
```

If we want to know what the population size is at the next time step, we can simply multiply N0 by (1+r).
```{r}
N0 * (1+r)
```

How can we find the population size for the next 50 years? Let's first make an object that is a vector of years using the **seq()** or "sequence" function.
```{r}
years <- seq(from=0, to=nyears, by=1)
years
```

Now, let's build a storage structure to store simulated population size over this time period

```{r}
N <- numeric(nyears+1)    # make an empty storage vector
N
```


### Run the simulation!

Then we can use a **for loop** (a very powerful computer programming trick) to automatically generate the population size for each of those years (note the similarity in the equation inside the for loop to Expression 1.15 in Gotelli).
```{r}
N[1] <- N0
lambda <- 1 + r           # (1 + r) is equal to lambda, the finite rate of growth 
for (i in 2:(nyears+1)){
  N[i] <- lambda*N[i-1]
}
N
```


### Plotting

Let's plot our population size against time. We only want the first 100 data points in N, so we use **indexing** to specify that. You may have noticed the square brackets inside the for loop...

```{r}
plot(N~years)
```

### Exercise 2 (R-related problems)

Please provide short answers to the following questions, and provide your R code to back up your answers.

Tweak the above code to run for 100 years. Plot your results. What is the final population size? 

Change r to 0.5 and run again for 100 years (and plot the results). What is the final population size now?

Try to tweak the value of $r$ such that the final population size after 100 years is 1000. What is the value of $r$? After you solve it by trial and error, can you solve this problem analytically using Eq. 10 above? 

Change the value of $r$ to -0.1. How long until the population goes extinct? Explain your answer. 

## Exponential growth in InsightMaker



































 

































